Multilevel generalizations of collective spin dynamics and squeezing

### 2020-07-22 collective_sun/multilevel_spin_notes.tex
minor changes in some conventions

Unlike the GGM operators in \eqref{eq:GGM_diag} and
\eqref{eq:GGM_off_diag}, the transition operators in
\eqref{eq:trans_ops} are not self-adjoint.  In order to express
Hamiltonians in terms of self-adjoint operators, we define the {\it
  drive operators}
\begin{align}
  D_{LM} \equiv \f{\eta_M}{\sqrt{2}}
  \sp{T_{LM} + \sign\p{M} T_{LM}^\dag},
  &&
  \eta_M \equiv
  \begin{cases}
    \sqrt{2} & M = 0 \\
    \p{-1}^M \sign\p{M}^{M+1/2} & M \ne 0
  \end{cases},
  \label{eq:drive_ops}
\end{align}
for integer $M$ with $\abs{M}\le L$.  The drive operator $D_{LM}$ is
proportional to the Hamiltonian induced on a nuclear spin by an
external classical field of spin-$L$ particles with spin projection
$M$ onto a quantization axis.  Similarly to the GGM operators in
\eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}, the drive operators
are proportional to the Pauli operators in the case of $n=2$, and
satisfy the orthonormality condition
\begin{align}
  \obk{D_{LM}|D_{L'M'}} = \delta_{LL'}\delta_{MM'},
\end{align}
which implies
\begin{align}
  H_{\t{int}} = -\f{u}{2} \sum_{p,q} \v D^{(p)}\c\v D^{(q)}
  = -\f{u}{2} \v\D \c \v\D,
  &&
  \v\D \equiv \sum_p \v D^{(p)},
  \label{eq:H_int_drive}
\end{align}
where $\v D$ is a vector of all drive operators in
\eqref{eq:drive_ops}.

@@

In particular, the amplitudes $\Omega_{1,0}$, $\Omega_{2,0}$, and
$\Omega_{2,1}$ can be simultaneously tuned to zero by setting
$\Omega_0=\Omega_+=\Omega_-=\xi_1\Omega$, at which point the Raman
drive becomes (up to an overall constant)

@@

$\Omega_0=\pm\Omega_\pm=\xi_1\Omega$, then up to a scalar term the
drive becomes

@@

We now consider the effect of applying periodic drive of the form
\begin{align}
  H_{\t{pd}}\p{t} = \sum_p \chi_p \omega \cos\p{\omega t} D_\z^{(p)},
\end{align}
where $\omega$ is a driving frequency, $\chi_p$ is a dimensionless
driving amplitude for spin $p$, and
$D_\z^{(p)} \equiv \xi_1 D_{1,0}^{(p)}$ with $\xi_1$ defined in
\eqref{eq:scale_fac}.  The periodic drive $H_{\t{pd}}\p{t}$ can be
implemented directly using external magnetic fields or driving lasers.
Transition operators $T_{LM}$ transform under rotations generated by
the drive operator $D_\z$ as
\begin{align}
  e^{i\phi D_\z} T_{LM} e^{-i\phi D_\z} = e^{iM\phi} T_{LM}.
\end{align}
If we move into the rotating frame of the drive, then the interaction
Hamiltonian transforms as
\begin{align}
  H_{\t{int}} \to \tilde H_{\t{int}}
  \equiv U_{\t{pd}}\p{t}^\dag H_{\t{int}} U_{\t{pd}}\p{t}
\end{align}
where
\begin{align}
  U_{\t{pd}}\p{t} \equiv \exp\sp{-i\int_0^td\tau\,H_{\t{pd}}\p{\tau}}
  = \prod_p \exp\sp{-i \chi_p \sin\p{\omega t} D_\z^{(p)}}.
\end{align}
Expanding the interaction Hamiltonian $H_{\t{int}}$ in terms of
transition operators $T_{LM}$, we find that
\begin{align}
  \tilde H_{\t{int}}
  = -\f{u}{2} \sum_{p,q,L,M} e^{iM\p{\chi_q-\chi_p}\sin\p{\omega t}}
  T_{LM}^{(p)\dag} T_{LM}^{(q)}
  = -\f{u}{2} \sum_{p,q,L,M,n}
  \J_n\p{M\sp{\chi_q-\chi_p}} e^{in\omega t}
  T_{LM}^{(p)\dag} T_{LM}^{(q)},
\end{align}
where $\J_n$ is the $n$-th order Bessel function of the first kind.
If the drive frequency $\omega$ is much larger than the interaction
strength $u$, i.e.~$\omega\gg u$, then we can neglect oscillating
terms with $n\ne0$, arriving at
\begin{align}
  \tilde H_{\t{int}} \approx -\f{u}{2} \sum_{p,q,L,M}
  \J_0\p{M\sp{\chi_q-\chi_p}}
  T_{LM}^{(p)\dag} T_{LM}^{(q)}.
\end{align}

@@

The drive operators are defined by
\begin{align}
  D_{LM} \equiv \f{\eta_M}{\sqrt{2}}
  \sp{T_{LM} + \sign\p{M} T_{LM}^\dag},
  &&
  \eta_M \equiv
  \begin{cases}
    \sqrt{2} & M = 0 \\
    \p{-1}^M \sign\p{M}^{M+1/2} & M \ne 0
  \end{cases},
\end{align}
or, equivalently,
\begin{align}
  D_{LM} \equiv \f{\eta_M}{\sqrt{2}}
  \sum_{s\in\set{\pm1}} \epsilon_{Ms} T_{L,sM},
  &&
  \epsilon_{Ms} \equiv
  \begin{cases}
    1 & s = +1 \\
    \p{-1}^M \sign\p{M} & s = -1
  \end{cases},
\end{align}
which implies that
\begin{align}
  T_{LM} = \f{\tau_M}{\sqrt{2}}
  \sum_{s\in\set{\pm1}} \sqrt{s} D_{L,sM},
  &&
  \tau_M \equiv
  \begin{cases}
    \sqrt{2}/\p{1+i} & M = 0 \\
    \eta_M^* & M \ne 0
  \end{cases},
\end{align}
where $\eta_M^*$ is the complex conjugate of $\eta_M$.  In order to
evaluate the product of two drive operators, we can thus expand them
in terms of transition operators, evaluate the corresponding products
of transition operators (see Appendix \ref{sec:trans_prod}), and
finally expand the result in terms of drive operators.  The product of
two drive operators is thus
\begin{align}
  D_{\ell_1 m_1} D_{\ell_2 m_2}
  = \sum_{L,M} g_{\ell_1m_1;\ell_2m_2}^{LM} D_{LM}
  \label{eq:drive_prod}
\end{align}
where
\begin{align}
  g_{\ell_1m_1;\ell_2m_2}^{LM}
  &\equiv \f{\eta_{m_1}\eta_{m_2}}{2\sqrt2}
  \sum_{r,s_1,s_2\in\set{\pm1}} \epsilon_{m_1s_1} \epsilon_{m_2s_2}
  \times f_{\ell_1,s_1m_1;\ell_2,s_2m_2}^{L,rM}
  \times \tau_{s_1m_1+s_2m_2} \times \sqrt{r}.
\end{align}
Redefining $\p{s_1,s_2}\to\p{rs_1,rs_2}$ and using the facts that
\begin{align}
  f_{\ell_1,rs_1m_1;\ell_2,rs_2m_2}^{L,rM}
  = r^{\ell_1+\ell_2+L} f_{\ell_1,s_1m_1;\ell_2,s_2m_2}^{LM},
  &&
  \tau_{rM} \sqrt{r} &= \tau_M \, \sigma_{Mr} \, r^M,
\end{align}
with
\begin{align}
  \sigma_{Mr} \equiv
  \begin{cases}
    \sqrt{r} & M = 0 \\
    \sign\p{M}^{\frac{1-r}{2}} & M \ne 0
  \end{cases},
  &&
  0^0 \equiv 1,
\end{align}
we can simplify
\begin{align}
  g_{\ell_1m_1;\ell_2m_2}^{LM}
  = \f{\eta_{m_1}\eta_{m_2}\tau_M}{\sqrt2}
  \sum_{s_1,s_2\in\set{\pm1}} f_{\ell_1,s_1m_1;\ell_2,s_2m_2}^{LM}
  \times \f12 \sum_{r\in\set{\pm1}}
  \epsilon_{m_1,r s_1} \epsilon_{m_2,r s_2} \,
  \sigma_{Mr} \, r^{\ell_1+\ell_2+L+M}.
  \label{eq:drive_struct}
\end{align}
In the case that $m_1=0$, we can simplify
\begin{align}
  g_{\ell_1,0;\ell_2,m}^{LM}
  \stackrel{m\ne 0}{=} f_{\ell_1,0;\ell_2 m}^{Lm}
  \times \delta_{\abs{M},\abs{m}} \times
  \begin{cases}
    \sp{\sp{\ell_1+\ell_2+L~\t{is even}}} & M = +m, \\
    i \sp{\sp{\ell_1+\ell_2+L~\t{is odd}}} & M = -m,
  \end{cases}
\end{align}
\begin{align}
  g_{\ell_1,0;\ell_2,0}^{LM} = f_{\ell_1,0;\ell_2,0}^{LM}.
\end{align}
Here $\sp{\sp{X}}\equiv1$ if $X$ is true and $0$ otherwise.

>>>>

Unlike the GGM operators in \eqref{eq:GGM_diag} and
\eqref{eq:GGM_off_diag}, the transition operators in
\eqref{eq:trans_ops} are not self-adjoint.  In order to express
Hamiltonians in terms of self-adjoint operators, we define the {\it
  drive operators}
\begin{align}
  D_{LM} \equiv \f{\eta_M}{\sqrt{2}}
  \sp{T_{LM} + \sign\p{M} T_{LM}^\dag},
  &&
  \eta_M \equiv
  \begin{cases}
    \sqrt{2} & M = 0 \\
    \p{-1}^M & M > 0 \\
    i & M < 0
  \end{cases},
  \label{eq:drive_ops}
\end{align}
for integer $M$ with $\abs{M}\le L$.  The drive operator $D_{LM}$ is
proportional to the Hamiltonian induced on a nuclear spin by an
external classical field of spin-$L$ particles with spin projection
$M$ onto a quantization axis.  Similarly to the GGM operators in
\eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}, the drive operators
are proportional to the Pauli operators in the case of $n=2$, and
satisfy the orthonormality condition
\begin{align}
  \obk{D_{LM}|D_{L'M'}} = \delta_{LL'}\delta_{MM'},
\end{align}
which implies
\begin{align}
  H_{\t{int}} = -\f{u}{2} \sum_{p,q} \v D^{(p)}\c\v D^{(q)}
  = -\f{u}{2} \v\D \c \v\D,
  &&
  \v\D \equiv \sum_p \v D^{(p)},
  \label{eq:H_int_drive}
\end{align}
where $\v D$ is a vector of all drive operators in
\eqref{eq:drive_ops}.

@@

Note that $\D_{0,0}$ is proportional to the identity operator, so the
amplitude $\Omega_{0,0}$ can be neglected without consequence.  If
$\Omega_0=\Omega_+=\Omega_-=\xi_1\Omega$, then all remaining
amplitudes except $\Omega_{1,1}$ and $\Omega_{2,2}$ vanish, and the
Raman drive becomes

@@

$\Omega_0=\pm\Omega_\pm=\xi_1\Omega$, then all amplitudes except
$\Omega_{2,1}$ and $\Omega_{2,2}$ vanish, and the drive becomes

@@

We now consider the effect of applying periodic drive of the form
\begin{align}
  H_{\t{pd}}\p{t} = \sum_p \chi_p \omega \cos\p{\omega t} S_\z^{(p)},
\end{align}
where $\omega$ is a driving frequency, $\chi_p$ is a dimensionless
driving amplitude for spin $p$, and
$S_\z^{(p)} \equiv \xi_1 D_{1,0}^{(p)}$ with $\xi_1$ defined in
\eqref{eq:scale_fac}.  The periodic drive $H_{\t{pd}}\p{t}$ can be
implemented directly using external magnetic fields or driving lasers.
Transition operators $T_{LM}$ transform under rotations generated by
the drive operator $S_\z$ as
\begin{align}
  e^{i\phi S_\z} T_{LM} e^{-i\phi S_\z} = e^{iM\phi} T_{LM}.
\end{align}
If we move into the rotating frame of the drive, then the interaction
Hamiltonian transforms as
\begin{align}
  H_{\t{int}} \to \tilde H_{\t{int}}
  \equiv U_{\t{pd}}\p{t}^\dag H_{\t{int}} U_{\t{pd}}\p{t}
\end{align}
where
\begin{align}
  U_{\t{pd}}\p{t} \equiv \exp\sp{-i\int_0^td\tau\,H_{\t{pd}}\p{\tau}}
  = \prod_p \exp\sp{-i \chi_p \sin\p{\omega t} S_\z^{(p)}}.
\end{align}
Expanding the interaction Hamiltonian $H_{\t{int}}$ in terms of
transition operators $T_{LM}$, we find that
\begin{align}
  \tilde H_{\t{int}}
  = -\f{u}{2} \sum_{p,q,L,M} e^{iM\p{\chi_q-\chi_p}\sin\p{\omega t}}
  T_{LM}^{(p)\dag} T_{LM}^{(q)}
  = -\f{u}{2} \sum_{p,q,L,M,n}
  \J_n\p{M\sp{\chi_q-\chi_p}} e^{in\omega t}
  T_{LM}^{(p)\dag} T_{LM}^{(q)},
\end{align}
where $\J_n$ is the $n$-th order Bessel function of the first kind.
If the drive frequency $\omega$ is much larger than the interaction
strength $u$, i.e.~$\omega\gg u$, then we can neglect oscillating
terms with $n\ne0$, arriving at
\begin{align}
  \tilde H_{\t{int}} \approx -\f{u}{2} \sum_{p,q,L,M}
  \J_0\p{M\sp{\chi_q-\chi_p}}
  T_{LM}^{(p)\dag} T_{LM}^{(q)}.
\end{align}

@@

The drive operators are defined by
\begin{align}
  D_{LM} \equiv \f{\eta_M}{\sqrt{2}}
  \sp{T_{LM} + \sign\p{M} T_{LM}^\dag},
  &&
  \eta_M \equiv
  \begin{cases}
    \sqrt{2} & M = 0 \\
    \p{-1}^M & M > 0 \\
    i & M < 0
  \end{cases},
\end{align}
or, equivalently,
\begin{align}
  D_{LM} \equiv \f{\eta_M}{\sqrt{2}}
  \sum_{s\in\set{\pm1}} \epsilon_{Ms} T_{L,sM},
  &&
  \epsilon_{Ms} \equiv
  \begin{cases}
    1 & s = +1 \\
    \p{-1}^M \sign\p{M} & s = -1
  \end{cases},
\end{align}
which implies that
\begin{align}
  T_{LM} = \f{\tau_M}{\sqrt{2}}
  \sum_{s\in\set{\pm1}} \sqrt{s} D_{L,sM},
  &&
  \tau_M \equiv
  \begin{cases}
    \sqrt{2}/\p{1+i} & M = 0 \\
    \eta_M^* & M \ne 0
  \end{cases},
\end{align}
where $\eta_M^*$ is the complex conjugate of $\eta_M$.  In order to
evaluate the product of two drive operators, we can thus expand them
in terms of transition operators, evaluate the corresponding products
of transition operators (see Appendix \ref{sec:trans_prod}), and
finally expand the result in terms of drive operators.  The product of
two drive operators is thus
\begin{align}
  D_{\ell_1 m_1} D_{\ell_2 m_2}
  = \sum_{L,M} g_{\ell_1m_1;\ell_2m_2}^{LM} D_{LM}
  \label{eq:drive_prod}
\end{align}
where
\begin{align}
  g_{\ell_1m_1;\ell_2m_2}^{LM}
  &\equiv \f{\eta_{m_1}\eta_{m_2}}{2\sqrt2}
  \sum_{r,s_1,s_2\in\set{\pm1}} \epsilon_{m_1s_1} \epsilon_{m_2s_2}
  \times f_{\ell_1,s_1m_1;\ell_2,s_2m_2}^{L,rM}
  \times \tau_{s_1m_1+s_2m_2} \times \sqrt{r}.
\end{align}
Redefining $\p{s_1,s_2}\to\p{rs_1,rs_2}$ and using the facts that
\begin{align}
  f_{\ell_1,rs_1m_1;\ell_2,rs_2m_2}^{L,rM}
  = r^{\ell_1+\ell_2+L} f_{\ell_1,s_1m_1;\ell_2,s_2m_2}^{LM},
  &&
  \tau_{rM} \sqrt{r} &= \tau_M \, \sigma_{Mr} \, r^M,
\end{align}
with
\begin{align}
  \sigma_{Mr} \equiv
  \begin{cases}
    \sqrt{r} & M = 0 \\
    \sign\p{M}^{\frac{1-r}{2}} & M \ne 0
  \end{cases},
  &&
  0^0 \equiv 1,
\end{align}
we can simplify
\begin{align}
  g_{\ell_1m_1;\ell_2m_2}^{LM}
  = \f{\eta_{m_1}\eta_{m_2}\tau_M}{\sqrt2}
  \sum_{s_1,s_2\in\set{\pm1}} f_{\ell_1,s_1m_1;\ell_2,s_2m_2}^{LM}
  \times \f12 \sum_{r\in\set{\pm1}}
  \epsilon_{m_1,r s_1} \epsilon_{m_2,r s_2} \,
  \sigma_{Mr} \, r^{\ell_1+\ell_2+L+M}.
  \label{eq:drive_struct}
\end{align}
In the case that $m_1=0$, we can simplify
\begin{align}
  g_{\ell_1,0;\ell_2,m}^{LM}
  \stackrel{m\ne 0}{=} f_{\ell_1,0;\ell_2 m}^{Lm}
  \times \delta_{\abs{M},\abs{m}} \times
  \begin{cases}
    \sp{\sp{\ell_1+\ell_2+L~\t{is even}}} & M = +m, \\
    i \sp{\sp{\ell_1+\ell_2+L~\t{is odd}}} & M = -m,
  \end{cases}
\end{align}
\begin{align}
  g_{\ell_1,0;\ell_2,0}^{LM} = f_{\ell_1,0;\ell_2,0}^{LM}.
\end{align}
Here $\sp{\sp{X}}\equiv1$ if $X$ is true and $0$ otherwise.
